A new metric for automatic discovery of periodic patterns in time series

### 2021-02-28 paper/main.Rmd
intro of normalisation and methodology re-written

The distance measure $wpd$ between two cyclic granularities $A$ and $B$ is aimed to capture the strength of the structure by estimating the maximum within-group and between-group variations. The steps employed for computing the distance measure is summarized as follows:

@@

@@

2. Percentiles of $v_{jk}$ are computed and stored in $q_{jk}$. Repeat for all pairs of categories of the form $(a_{j} b_{k}, a_{j'}b_{k'}): j = 1, 2, \dots, J\}, B = \{ b_k: k = 1, 2, \dots, K\}$.

3. The pairwise distances between pairs $(a_{j} b_{k}, a_{j'}b_{k'})$  denoted by $d_{(jk, j'k')} = JSD(q_{jk}, q_{j'k'})$ is computed.

4. The pairwise distances $d_{(jk, j'k')}$ is transformed using a suitable tuning parameter ($0<\lambda<1$) depending on if they are within-facet($d_w$) or between-facets($d_b$) as follows:

@@

6. Repeat for all harmony pairs in $N_C$.

@@

A simulation study is carried out to explore how $wpd$ performs under various designs, its parameters and limitations. To this end, simulations were carried out for four different designs and the following factors that could potentially have an impact on the values of $wpd$:

- $nx$ (number of levels of x-axis)
- $nfacet$ (number of levels of facets)
- $\lambda$ (tuning parameter)
- $\omega$ (increment in each panel design)
- $dist$ (normal/non-normal distributions with different location and scale)
- $n$ (sample size for each combination of categories)
- $nsim$ (number of simulations)
- $nperm$ (number of permutations of data)  
- $designs$  
$D_{null}$  (No difference in distribution)  
$D_{var_f}$ (Difference in distribution only across facets)  
$D_{var_x}$ (Difference in distribution only across x-axis)  
$D_{var_{all}}$ (Difference in distribution in both facets and x-axis)

Results are presented in two parts. The dependence of $wpd$ on $nx$ and $nfacet$ under $D_{null}$ is presented here, which lays the foundation for the next section. The rest of the results that discusses the relationship of the $wpd$ with other factors is presented in the Supplementary section of the paper.

### Design of the simulation study

Observations are generated from a Gamma(2,1) distribution for each combination of $nx$ and $nfacet$ from the following sets: $nx = nfacet = \{2, 3, 5, 7, 14, 20, 31, 50\}$ to cover a wide range of levels from very low to moderately high. Each combination is being referred to as a _panel_. That is, data is being generated for each of the panels $\{nx = 2, nfacet = 2\}, \{nx = 2, nfacet = 3\}, \{nx = 2, nfacet = 5\},  \dots, \{nx = 50, nfacet = 31\}, \{nx = 50, nfacet = 50\}$. For each of the $64$ panels, $ntimes = 500$ observations are drawn for each combination of the categories. That is, if we consider the panel  $\{nx = 2, nfacet = 2\}$, $500$ observations are generated for each of the combination of categories from the panel, namely, $\{(1, 1), (1, 2), (2, 1), (2, 2)\}$. The values of $\lambda$ is set to $0.67$ and values of raw wpd $wpd$ is obtained. This entire design applies to the null cases and hence there is no difference in distribution between any categories in the panel.

@@

Figure \ref{fig:raw} shows the distribution of $wpd$ plotted across different nx and nfacet categories. Both shape and scale of the distributions change across panels. This is not desirable as it would mean we would not be able to compare $wpd$ across different $nx$ and $nfacet$ as each of them are drawn from distributions with different locations and scale. In Figure \ref{fig:quadratic}, we see how the median of $wpd_{raw}$ varies with the total number of distances $nx*nfacet$ for each panel. The median increases abruptly for lower values of $nx*nfacet$ and slowly for higher $nx*nfacet$.

```{r raw, fig.cap = "Distribution of raw wpd is plotted across different nx and nfacet categories. Both shape and scale of the distribution changes for different nx and nfacet categories."}
G21 <- read_rds("simulations/raw/null_design_quantrans/data-agg/all_data_wpd_Gamma21.rds")

G21 %>% 
  ggplot(aes(x = value)) + 
  geom_density(fill = "blue") +
  facet_grid(nx~nfacet,
             labeller = "label_both") + 
  scale_x_continuous(breaks = scales::breaks_extended(3)) + 
  xlab("raw wpd")
```

```{r quadratic, fig.cap = "$wpd_{raw}$ is plotted against nx*nfacet and the blue line represents the median of the multiple values for each nx*nfacet levels."}
G21 %>% 
  ggplot(aes(x=nx*nfacet, y = value)) +
  geom_point() + stat_summary(fun=median, geom="line", aes(group=1), color = "blue") + xlab("nx*nfacet") + ylab("raw wpd")

@@

The distribution of wpd is different for different levels of facets and x-axis levels. This is because the statistic maximum which is used to define $wpd$ is affected by the number of categories. The measure would have higher values if $A$ or $B$ has higher levels. However, we would ideally want a higher value of the measure only if there is significant difference between distributions across facet or x-axis categories, and not because the number of categories $J$ or $K$ is high. Therefore, in order to compare $wpd$ across different combinations of facet and x-axis levels, we need to eliminate the impact of different levels of the facets and x-axis first and get a normalized measure. Henceforth we call the measure already discussed as $wpd_{raw}$ and the normalized measure as $wpd_{norm}$. The measure $wpd_{norm}$ could potentially lead to comparison of the measure across different panels and also identifying only the interesting panels from a data set. We discuss two approaches for normalization, both of which are based on the simulation results.

@@

## Simulation environment

@@

We need a transformation on ${wpd}$ which will make it independent of the values of $nx*nfacet$. Two approaches have been employed for that purpose, the first one involves fitting a model and the latter involves a permutation method to make the distribution of the transformed $wpd$ similar across different $nx$ and $nfacet$.

@@

The permutation approach ensures that the distribution of the normalized distance measure has the same mean and standard deviation across all combinations of $nx_i$ and $nfacet_j$. The normalized distances through permutation is computed as follows: $x^{perm}_{k} =  (x_{k} - mean_l(x_{k,l}))/sd_l(x_{k, l})$,  $x^{perm}_{k}$ is the value of the $wpd_{perm}$ for the $k^{th}$ panel. Standardizing the variable $wpd_{perm}$ in this approach leads to $location = 0$ and $scale = 1$ for this variable.

While this works successfully to make the mean and standard deviation across different $nx$ and $nfacet$ (as seen in Figure \ref{fig:}), it is computationally heavy and time consuming, and hence less user friendly when being actually used in practice. Hence, we propose another approach to normalization which is more approximate than exact but still has the same accuracy when compared to the permutation approach.

@@

>>>>

The distance measure $wpd$ between two cyclic granularities $A$ and $B$ is aimed to capture the strength of the structure by estimating the maximum within-group and between-group variations. Furthermore, the intended aim of $wpd$ is to capture differences in categories irrespective of the distribution from which the data is generated. Hence, as a pre-processing step,the raw data is normal quantile transformed so that the quantile of the transformed data follows a standard normal distribution. The steps employed for computing the distance measure is summarized as follows:

<!-- The values of $wpd$ indeed depend upon the underlying distribution. This is also not desirable as the intended aim of $wpd$ is to capture differences in categories irrespective of the distribution from which the data is generated. The steps employed for computing the distance measure is summarized as follows: -->

@@

1. Perform NQT on the measured variable $v_t$ to obtain $v^*_t$.

2. Fix harmony pair $(A, B)$.

@@

3. Percentiles of $v^*_{jk}$ are computed and stored in $q_{jk}$. Repeat for all pairs of categories of the form $(a_{j} b_{k}, a_{j'}b_{k'}): j = 1, 2, \dots, J\}, B = \{ b_k: k = 1, 2, \dots, K\}$.

4. The pairwise distances between pairs $(a_{j} b_{k}, a_{j'}b_{k'})$  denoted by $d_{(jk, j'k')} = JSD(q_{jk}, q_{j'k'})$ is computed.

5. The pairwise distances $d_{(jk, j'k')}$ is transformed using a suitable tuning parameter ($0<\lambda<1$) depending on if they are within-facet($d_w$) or between-facets($d_b$) as follows:

@@

6. Repeat Steps 2-5 for all harmony pairs in $N_C$.

@@

Simulations were carried out to explore the behavior of $wpd$ under the following factors that could potentially impact the values of $wpd$:

- $nx$ (number of levels of x-axis)
- $nfacet$ (number of levels of facets)
- $\lambda$ (tuning parameter)
- $\omega$ (increment in each panel design)
- $dist$ (normal/non-normal distributions with different location and scale)
- $n$ (sample size for each combination of categories)
- $designs$ ($D_{null}$, $D_{var_f}$, $D_{var_x}$ and $D_{var_{all}}$)

<!-- - $nsim$ (number of simulations) -->
<!-- - $nperm$ (number of permutations of data)   -->

Results are presented in two parts. The dependence of $wpd$ on $nx$ and $nfacet$ under $D_{null}$ is presented here, which lays the foundation for the next section. The rest of the results that discusses the relationship of the $wpd$ with other factors is presented in details in the Supplementary section of the paper. They show that the designs $D_{var_f}$ and $D_{var_x}$ intersect at $\lambda = 0.5$ and hence for up-weighing designs of the form $D_{var_x}$, $\lambda = 0.67$ has been considered for computation of $wpd$ in the rest of the paper.

### Simulation design

Observations are generated from a Gamma(2,1) distribution for each combination of $nx$ and $nfacet$ from the following sets: $nx = nfacet = \{2, 3, 5, 7, 14, 20, 31, 50\}$ to cover a wide range of levels from very low to moderately high. Each combination is being referred to as a _panel_. That is, data is being generated for each of the panels $\{nx = 2, nfacet = 2\}, \{nx = 2, nfacet = 3\}, \{nx = 2, nfacet = 5\},  \dots, \{nx = 50, nfacet = 31\}, \{nx = 50, nfacet = 50\}$. For each of the $64$ panels, $ntimes = 500$ observations are drawn for each combination of the categories. That is, if we consider the panel  $\{nx = 2, nfacet = 2\}$, $500$ observations are generated for each of the combination of categories from the panel, namely, $\{(1, 1), (1, 2), (2, 1), (2, 2)\}$. The values
of $wpd$ is obtained for each of the panels. This design corresponds to $D_{null}$ as each combination of categories in a panel are drawn from the same distribution. Furthermore, the data is simulated for each of the panels $nsim=200$ times, so that the distribution of $wpd$ under $D_{null}$ could be observed.

@@

Figure \ref{fig:raw} shows the distribution of $wpd$ plotted across different $nx$ and $nfacet$ categories. Since under $D_{null}$, there is no difference in distributions across different categories, we expect the distance measure $wpd$ to reflect that as well and have the same null distribution across categories. But Figure \ref{fig:raw} shows that both the location and scale of the distributions change across panels. This is not desirable under $D_{null}$ as it would mean comparisons of $wpd$ values is not appropriate across different $nx$ and $nfacet$. Figure \ref{fig:quadratic} shows how the median of $wpd_{raw}$ varies with the total number of distances $nx*nfacet$ for each panel. The median increases abruptly for lower values of $nx*nfacet$ and slowly for higher $nx*nfacet$.

<!-- Furthermore, simulation results also show that the values of $wpd$ indeed depend upon the underlying distribution. This is also not desirable as the intended aim of $wpd$ is to capture differences in categories irrespective of the distribution from which the data is generated. -->

```{r raw, fig.cap = "Distribution of $wpd$ is plotted across different $nx$ and $nfacet$ categories. Both shape and scale of the distribution changes for different nx and nfacet categories."}
G21 <- read_rds("simulations/raw/null_design_quantrans/data-agg/all_data_wpd_Gamma21.rds")

G21 %>% 
  ggplot(aes(x = value)) + 
  geom_density(fill = "blue") +
  facet_grid(nx~nfacet,
             labeller = "label_both") + 
  scale_x_continuous(breaks = scales::breaks_extended(3)) + 
  xlab("wpd")
```

```{r quadratic, fig.cap = "$wpd$ is plotted against $nx*nfacet$ (the maximum number of pairwise comparisons) and the blue line represents the median of the multiple values for each $nx*nfacet$ levels. The median increases abruptly for lower values of $nx*nfacet$ and slowly for higher $nx*nfacet$."}
G21 %>% 
  ggplot(aes(x=nx*nfacet, y = value)) +
  geom_point() + stat_summary(fun=median, geom="line", aes(group=1), color = "blue") + xlab("nx*nfacet") + ylab("wpd")

@@

The distribution of $wpd$ is different for different levels of facets and x-axis levels. This is because the statistics maximum which is used to define $wpd$ is affected by the number of comparisons (resulting pairwise distances). The measure would have higher values if $A$ or $B$ has higher levels. However, we would ideally want a higher value of the measure only if there is a significant difference between distributions across facet or x-axis categories, and not because the number of categories $J$ or $K$ is high. Therefore, in order to compare $wpd$ across different combinations of facet and x-axis levels, we need to eliminate the impact of different number of comparisons and get a normalized measure. Henceforth, we call the normalized measure as $wpd_{norm}$. The measure $wpd_{norm}$ could potentially lead to comparison of the measure across different panels and also help distinguishing the interesting panels from a data set. We discuss two approaches for normalization, both of which are based on the simulation results.

@@

@@

We need a transformation on ${wpd}$ which will make it independent of the values of $nx*nfacet$. Two approaches have been employed for that purpose, the first one involves fitting a model and the latter involves a permutation method to make the distribution of the transformed $wpd$ similar for different comparisons.

@@

This method is somewhat similar in spirit to bootstrap or permutation tests, where the goal is to test the hypothesis that the groups under study have identical distributions. This method, in essence, accomplishes a different goal of making the location and scale of different groups (panels) same under $D_{null}$. The steps are as follows:

1. Compute $wpd$ for a harmony pair (A, B) for the original measured variable $v_t$ and store it in $wpd^{orig}$.

2. Consider a permutation of the original measured variable $v_t^{perm_1}$ and again compute $wpd$ for the permuted data. Store it in $wpd^{perm_1}$.

3. Repeat Step 2 for a large number ($nperm = 200$) of random permutations of the data yielding $nperm$ values : $wpd^{perm_1}, wpd^{perm_2}, \dots, wpd^{perm_{nperm}}$. Store the vector in $wpd^{perm}$.

4. Define $wpd^{perm}_{norm} = \frac{(wpd^{orig} - \bar{wpd^{perm}})}{sd(wpd^{perm})}$, where $\bar{wpd^{perm}}$ and $sd(wpd^{perm})$ are the mean and standard deviation of $wpd^{perm}$ respectively.

<!-- Let $\{x_{k,l}, k = 1, 2, \dots, nx*nfacet, l = 1, 2, \dots, nsim\}$ denote the values of $wpd$ obtained from the simulation study for $k^{th}$ panel in the $i^{th}$ simulation. Hence, for each of those $k$ panel, we have $nsim$ values of $wpd$.combinations of $nx_i$ and $nfacet_j$. The normalized distances through permutation is computed as follows: $x^{perm}_{k} =  (x_{k} - mean_l(x_{k,l}))/sd_l(x_{k, l})$,  $x^{perm}_{k}$ is the value of the $wpd_{perm}$ for the $k^{th}$ panel-->

Standardizing the variable $wpd$ in the permutation approach ensures that the distribution of $wpd^{perm}_{norm}$ has the same $mean = 0$ and $sd = 1$ across all comparisons under $D_{null}$. While this works successfully to make the location and scale similar across different $nx$ and $nfacet$ (as seen in Figure \ref{fig:norm}), it is computationally heavy and time consuming, and hence less user friendly when being actually used in practice. Hence, we propose another approach to normalization which is more approximate than exact but still has the same accuracy when compared to the permutation approach.

```{r norm, fig.cap = "Distribution of $wpd^{perm}_{norm}$ is plotted across different $nx$ and $nfacet$ categories. Both shape and scale of the distributions are now similar for different panels under the null design."}
G21_norm <- read_rds(here("simulations/norm/null_design_quantrans_nperm/data-agg/all_data_wpd_Gamma21.rds"))

G21_norm %>% 
  ggplot(aes(x = value)) + 
  geom_density(fill = "blue") +
  facet_grid(nx~nfacet,
             labeller = "label_both") + 
  scale_x_continuous(breaks = scales::breaks_extended(3)) + 
  xlab("wpd normalised using permutation approach")
```

@@

# Simulation environment
